---
title: "Exploring the One Earth Bioregions Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the One Earth Bioregions Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4, dpi = 100)
```

```{r setup}
library(oneearthr)
library(terra)
```

The One Earth Bioregions Framework organises the terrestrial realm into a
four-level hierarchy:

| Level | Features | Function |
|-------|----------|----------|
| Realm | 14 | `oe_layer("realms")` |
| Subrealm | 53 | `oe_layer("subrealms")` |
| Bioregion | 185 | `oe_layer("bioregions")` |
| Ecoregion | 847 | `oe_layer("ecoregions")` |

## The layers

```{r realms}
realms <- oe_layer("realms")
plot(realms, col = as.factor(realms$biogeorelm), border = "grey40",
     main = "Biogeographical realms")
```

```{r subrealms}
subrealms <- oe_layer("subrealms")
plot(subrealms, col = as.factor(subrealms$biogeorelm), border = "grey60",
     main = "Subrealms")
```

```{r bioregions}
bioregions <- oe_layer("bioregions")
plot(bioregions, col = as.factor(substr(bioregions$bioregions, 1, 2)),
     border = "grey70", main = "Bioregions (coloured by realm prefix)")
```

```{r ecoregions}
ecoregions <- oe_layer("ecoregions")   # simplified geometry
plot(ecoregions, col = as.factor(ecoregions$realm), border = NA,
     main = "RESOLVE Ecoregions 2017 (simplified)")
```

For publication-quality boundaries, download the original geometries once;
they are cached per user. `scale = "large"` works for all four layers — the
ecoregions (~150 MB) come from the public RESOLVE bucket, the framework
layers from the One Earth shared folders:

```r
ecoregions_full <- oe_layer("ecoregions", scale = "large")
realms_full     <- oe_layer("realms", scale = "large")
```

## Joining levels

The bioregion layer only carries codes (e.g. `AN03`), so the hierarchy is
best traversed spatially. A spatial join from ecoregion centroids to the
coarser levels recovers the full path for any ecoregion:

```{r join}
pts <- centroids(ecoregions)

sub_of <- terra::relate(pts, subrealms, relation = "intersects")
sub_idx <- apply(sub_of, 1, function(x) which(x)[1])
ecoregions$subrealm <- subrealms$subrealm_n[sub_idx]

head(ecoregions[, c("eco_name", "realm", "subrealm")] |> values())
```

And the reverse — listing every ecoregion inside one subrealm:

```{r filter}
sa_grass <- ecoregions[ecoregions$subrealm == "South America Grasslands", ]
plot(sa_grass, col = "darkolivegreen3",
     main = sprintf("%d ecoregions in South America Grasslands",
                    nrow(sa_grass)))
```

### Bioregion names

The bioregion geometry only carries codes; `oe_bioregion_names` maps them
to names and their position in the hierarchy:

```{r names}
head(oe_bioregion_names)

bioregions <- merge(bioregions, oe_bioregion_names,
                    by.x = "bioregions", by.y = "code")
bioregions[bioregions$name == "Amazon River Estuary", ]
```

## Attribute reference (ecoregions)

The ecoregion layer keeps the original RESOLVE attributes, including the
protection-status fields of the framework:

```{r attrs}
table(ecoregions$nnh_name)
```

`nnh_name` ("Nature Needs Half" assessment) classifies each ecoregion as
*Protected*, *Half Protected*, *Nature Could Reach Half Protected*, or
*Nature Imperiled* — the colour ramps `color`, `color_bio`, `color_nnh`
ship with the data for reproduction of the official maps:

```{r colors}
plot(ecoregions, col = ecoregions$color_nnh, border = NA,
     main = "Ecoregion conservation status (official colours)")
```

## Data sources

- Dinerstein et al. (2017), *BioScience* 67(6),
  [doi:10.1093/biosci/bix014](https://doi.org/10.1093/biosci/bix014)
  (CC-BY 4.0) — ecoregions.
- One Earth Bioregions Framework 2023, <https://www.oneearth.org> —
  realms, subrealms, bioregions.
